##REBASE R notebook - 

####Reading in the data

(From Coursera R Programming-Week 1 - Reading large tables)
Help R read the file by specifying beforehand: how many rows, column classes
Take an initial sample of REBASE csv file (enz.csv)
```{r}
initial<-read.csv("C:\\Users\\Kelly\\Documents\\Python\\REBASE\\enz.csv",nrows=100,header=T)
print(initial[1:2,])
```
```{r}
classes<-sapply(initial,class)
print(classes)
```
Now read table using these variables.
```{r}
tabAll<-read.table("C:\\Users\\Kelly\\Documents\\Python\\REBASE\\enz.csv", nrow=20, colClasses = classes,header=T, sep=",")
print(tabAll[1:3,])
```

Calculate memory requirements:
enz.csv has 11569 rows, 5 columns, 835kb csv file
11569 x 5 x 8bytes/numeric value
```{r}
bytes= 11569 * 5 * 8
print(bytes)
```

```{r}
MB = bytes / 2^20
GB = MB/1000
print(data.frame(c("MB=","GB="),c(MB,GB)))
```
Need twice the amount of MB calculated to actually read.
=MB * 2 = 0.883 MB (883 kb)

For ex:
if dataset is 1.34 GB and computer has 2 GB of RAM, it will not be able to read in this dataset.


